Verified Nanopasses for Compiling Conditionals

### 2025-08-12 paper.tex
edits

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents an elegant family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that (1) no longer
contains Boolean expressions, (2) the condition expression of a
condition is a value, and (3) jumps are expressed using zero-arity
function calls. One of the key parts of that translation is in the
compilation of a conditional expression such as $(\IF{b}{e_1}{e_2})$,
where the expression $b$ is compiled using a special function
$\mathcal{B}$ that takes two more arguments, the generated code for
the two branches, that is, for the two continuations of $b$. Thus,
$\mathcal{B}$ can analyze the expression $b$ to decide how to generate
code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is even
necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}
\noindent The treatment code as continuation goes back to earlier work
by \citet{Lawall:1994aa} on continuation-based partial
evaluation. Indeed, the above is a partial evaluation of Boolean
expressions.

@@

The CompCert compiler \citep{Leroy:2006fe} (a compiler for C verified
in Rocq) uses the one-function per context approach of Danvy in the
RTLgen pass that translates from the CminorSel intermediate language
to CompCert's register transfer language. The verified CakeML
compiler, on the other hand, employs a naive translation of Boolean
expressions and conditionals for the compilation of the stackLang 
intermediate language into labLang~\citep{Kumar:2014aa}.

Towards developing a course on verified compilation, I am interested
in adapting these verification techniques to the compiler
presented in the \emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. That book takes an
incremental approach to teaching compilers, that is, it guides the
students through the creation of multiple compilers for a sequence of
languages, each of which adds a language feature to the previous
language. For the language with Boolean expressions and conditionals,
the Explicate Control pass of the compiler uses the one-function per
context approach of Danvy to translate into an x86-flavored register
transfer language.  \emph{Essentials of Compilation} also employs the
nanopass approach, breaking up the compiler into many small passes
that are easier for student to digest~\citep{Sarkar:2004fk}.  There
are many other pedogogical treatments of
compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa};
what makes the \emph{Essentials of Compilation} unique is that it is
the first textbook to apply the incremental approach.  However,
\emph{Essentials of Compilation} does not include material about the
formal verification of compilers; hence my present interest.

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. The Agda code is available in the supplementary
material. It does not depend on particular features of Agda and could
have been carried out in Rocq, Isabelle, or Lean.\footnote{The choice
of Agda for this paper is not an endorsement of Agda but is a
byproduct of the author's relative expertise in Agda at the time of
writing.}

@@

Section~\ref{sec:langs} of the paper introduces the intermediate
languages. Section~\ref{sec:compiler} describes the compiler,
detailing each pass and the proof of correctness.

@@

In this section we describe how we formalize the syntax and semantics
of the source language and four intermediate languages in Agda:
\begin{enumerate}
\item Integers, Booleans, \texttt{let}, and \texttt{if}
\item Monadic Normal Form
\item Imperative with Structured Control Flow
\item A C-like language with \emph{goto}
\item X86 assembly but with variables
\end{enumerate}
These languages and their semantics are quite standard. However, there
are many other intermediate languages in the literature to choose
from~\citep{Flanagan:1993cg,Hatcliff:1994vn,Kelsey:1995uq,Birkedal:1996aa},
so why these languages, and why in this particular order?  The answer
is that they happen to represent the images of the four nanopasses.
So then the next question is: why do we order the nanopasses in this
way? The main thing to consider is that some passes make later passes
easier because they remove a language feature that would have been
complicated to handle. For example, the Lift Locals pass removes
\texttt{let}, and I found \texttt{let} difficult to deal with in the
proof of correctness for Explicate Control (\texttt{let} is relatively
easy to handle in the code for Explicate Control). The other thing to
consider is that some compiler passes use certain language features in
their generated code, so the passes that remove those language
features better come later in the ordering.  For example, the Closure
Conversion pass (not discussed here) generates code with \texttt{let},
so Closure Conversion comes before Lift Locals. Finally, many design
choices in \emph{Essentials of Compilation} have been inherited from
\citet{Dybvig:2010aa}

One of the perennial challenges of formalizing semantics in a proof
assistant is that one needs to choose a representation of variables.
Of the many options, my current favorite is good old de Bruijn
indices~\citep{Bruijn:1972kx}.  Thus, identifiers \lstinline{Id} are
represented as natural numbers.

\begin{lstlisting}
Id = $ℕ$
\end{lstlisting}

@@

\section{Conclusion}

>>>>

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents an elegant family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that (1) no longer
contains Boolean expressions, (2) the condition expression of a
condition is a value, and (3) jumps are expressed using zero-arity
function calls. One of the key parts of that translation is in the
compilation of a conditional expression such as $(\IF{b}{e_1}{e_2})$,
where the expression $b$ is compiled using a special function
$\mathcal{B}$ that takes two more arguments, the generated code for
the two branches, that is, for the two continuations of $b$. Thus,
$\mathcal{B}$ can analyze the expression $b$ to decide how to generate
code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is even
necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}
\noindent This idea builds on earlier work by \citet{Lawall:1994aa} on
continuation-based partial evaluation. Indeed, the above is a partial
evaluation of Boolean expressions.

@@

The CompCert compiler \citep{Leroy:2006fe} (a compiler for C verified
in Rocq) uses the one-function per context approach of Danvy in the
RTLgen pass that translates from the CminorSel intermediate language
to CompCert's register transfer language. The verified CakeML
compiler, on the other hand, employs a naive translation of Boolean
expressions and conditionals for the compilation of stackLang into
labLang~\citep{Kumar:2014aa}.

Towards developing a course on verified compilation, I am interested
in proving correctness of the compiler presented in the
\emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. While it does not include
proofs of correctness for its compiler passes, the textbook is
organized in a way that might facilitate the proofs of correctness.
%
\emph{Essentials of Compilation} takes an \emph{incremental} approach
to teaching compilers, that is, it guides the students through the
creation of multiple compilers for a sequence of languages, each of
which adds a language feature to the previous language. For the
language with Boolean expressions and conditionals, the Explicate
Control pass of the compiler uses the one-function per context
approach of Danvy to translate into an x86-flavored register transfer
language.
%
In addition to being incremental, \emph{Essentials of Compilation}
employs the \emph{nanopass} approach, breaking up the compiler into
many small passes that are easier for students to
digest~\citep{Sarkar:2004fk}.
%
There are many other pedogogical treatments of
compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa}
but none of them take the incremental approach.

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. The Agda code is available in the supplementary
material and at the following location.
\begin{center}
\url{https://github.com/jsiek/correct_compilers}
\end{center}
The Agda code does not depend on special features of Agda and could
have been carried out in Rocq, Isabelle, or Lean.\footnote{The choice
of Agda for this paper is not an endorsement of Agda but is a
byproduct of the author's relative expertise in Agda at the time of
writing.}

@@

Section~\ref{sec:langs} of the paper introduces the intermediate
languages. Section~\ref{sec:compiler} describes the compiler,
detailing each pass and the proof of
correctness. Section~\ref{sec:conclusion} concludes the paper.

@@

In this section we describe how we formalize the syntax and semantics
of the source language and four intermediate languages in Agda:
\begin{enumerate}
\item Integers, Booleans, \texttt{let}, and \texttt{if}
\item Monadic Normal Form
\item Imperative with Structured Control Flow
\item A C-like language with \emph{goto}
\item X86 assembly but with variables
\end{enumerate}
These languages and their semantics are quite standard. However, there
are many other intermediate languages in the literature to choose
from~\citep{Flanagan:1993cg,Hatcliff:1994vn,Kelsey:1995uq,Birkedal:1996aa},
so why these languages, and why this order?  The answer is that they
happen to represent the images of the four nanopasses.  So then the
next question is: why do we order the nanopasses this way? The main
thing to consider is that some passes make later passes easier because
they remove a language feature that would have been complicated to
handle. For example, the Lift Locals pass removes \texttt{let}, and I
found \texttt{let} difficult to deal with in the proof of correctness
for Explicate Control. The other thing to consider is that some
compiler passes use certain language features in their generated code,
so the passes that remove those language features better come later in
the ordering.  For example, the Closure Conversion pass (not discussed
here) generates code with \texttt{let}, so Closure Conversion comes
before Lift Locals. Finally, many design choices in \emph{Essentials
of Compilation} have been inherited from \citet{Dybvig:2010aa}

One of the perennial challenges of formalizing semantics in a proof
assistant is that one needs to choose a representation of variables.
Of the many options, my current favorite is good old de Bruijn
indices~\citep{Bruijn:1972kx}.  Thus, identifiers \lstinline{Id} are
represented as natural numbers.
\begin{lstlisting}
Id = $ℕ$
\end{lstlisting}

@@

\section{Conclusion}
\label{sec:conclusion}
